---
title: "VegX"
output: github_document
---

```{r setup, echo = FALSE}

knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.path = "README-")
```

## Introduction

A primary technical impediment to large-scale sharing of vegetation data is the lack of a recognized international exchange standard for linking the panoply of tools and database implementations that exist among various organizations and individuals participating in vegetation research. 

## The Veg-X exchange standard

The **Veg-X exchange standard** for plot-based vegetation data (Wiser et al. 2011) is intended to be used to share and merge vegetation-plot data of different kinds. Veg-X allows for observations of vegetation at both individual plant and aggregated observation levels. It ensures that observations are fixed to physical sample plots at specific points in space and time, and makes a distinction between the entity of interest (e.g. an individual tree) and the observational act (i.e. a measurement). The standard supports repeated measurements of both individual organisms and plots, and enables the connection between the entity observed and the taxonomic concept associated with that observation to be maintained.


## R package to use the Veg-X standard

Functions to import, integrate, harmonize and export vegetation data using the VegX standard (ver. 2.0.0)

### Installation

```{r installation, eval = FALSE}
devtools::install_github("miquelcaceres/VegX", build_vignettes=TRUE)
```


## References

+ Wiser SK, Spencer N, De Caceres M, Kleikamp M, Boyle B & Peet RK (2011). Veg-X - an exchange standard for plot-based vegetation data. *Journal of Vegetation Science* 22: 598-609.

