---
title: "Using the VegX package"
author: "IAVS Ecoinformatics Working Group"
date: "`r Sys.Date()`"
output: 
  rmarkdown::html_vignette:
    css: mystyle.css
vignette: >
  %\VignetteIndexEntry{Veg-X standard documentation}
  %\VignetteEngine{knitr::rmarkdown}
  \usepackage[utf8]{inputenc}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## About this manual

In this vignette you will learn how to use the **VegX R package** to integrate and harmonize vegetation data. For the examples, we use the example data sets provided in the package. If you do not know what Veg-X is, please refer to vignette *The Veg-X standard and the VegX R package*.

```{r load VegX}
library(VegX)
```

## Creating and populating Veg-X documents

### Loading the source data into R

```{r load source data}
data(mokihinui)
```


### Creating a new Veg-X document
```{r init}
x = newVegX()
```

### Adding project, plot and observation dates

You can also embed plots, for example:

```{r add plot obs}
# Define mapping
mapping = list(projectTitle = "Project", plotName = "Plot", subPlotName = "Subplot",
               obsStartDate = "PlotObsStartDate", obsEndDate = "PlotObsStopDate")

# Adds information about projects, plots and plot observations (no measurements yet)
x = addPlotObservations(x, site, mapping = mapping)
```

```{r summary}
summary(x)
```

```{r show plot obs}
head(showElementTable(x, "plotObservation"),10)
```
